---
title: "Non Linear Recovery"
author: "David Champredon"
date: "August 25, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Background

In a standard SIR model, the number of individuals that recover from the infection is linear withrespect to time, that is the ODE is:
$$\frac{dI}{dt} = \beta SI - \gamma I $$

The variable $I$ is the prevalence, expressed as a proportion of the total population (hence between 0 and 1). We want to understand the effect of a non-linear recovery on both the infectious duration distribution and also on the generation interval distributions. 

We write the new dynamical system as 
$$ \frac{dI}{dt} = \beta SI - \gamma K(I) I $$
where $K$ is a non-negative continous function on [0;1].

See DE's notes for an analytical exploration. Here we focus on a numerical exploration.


## Epidemiological Model

An agent-based model is used to explore the effect of non-linear recovery. The model is a stochastic $SE_mI_nR$ compartmental model implemented with a tau-leap algorithm (there is the possibility to run with an exact Gillespie algorithm, but it is very slow). The only information that is tracked at the indivdual level is the time of infection, the time of disease transmission to another individual, the time infectiousness starts and ends. The first two times give the generation interval distributions, the last two the infectious duration one.

The code is written in C++ and wrapped in a R library. 


## Implemented $K$ functions

For now, six functions are implemented in the code. 

Function name  | Formula
------------- | -------------
`one`  | $K(I) = 1$
`lin` | $K(I) = I$
`sqrt` | $K(I) = \sqrt{I}$
`exp` | $K(I) = \exp{\left(-aI\right)}$
`inv` | $K(I) = 1 / (a+bI)$
`pow` | $K(I) = I^{a-1}$

Some names are redundant (e.g. `lin` could be expressed as `pow`) but that makes labelling easier.


## Basic consistency checks

First of all, let's check there are no obvious implementation errors. The ABM with $K(I)=1$ is compared with both an ODE SIR model solved with `lsoda` and a stochastic SIR model using the `adaptivetau` package.

```{r ode-stoch-abm-checks, echo=FALSE, cache=TRUE, fig.height=10, message=FALSE, warning=FALSE,results='hide'}
libpath <-  "./ABM/Rlibrary/lib/"
library(nlr,lib.loc =libpath)
source('./SIR_nlr_ODE.R')
source('./SIR_nlr_stoch.R')
source('./SIR_nlr_ABM.R')

save.to.file <- FALSE

### ==== Shared parameters ====
horizon = 1
pop.size <- 2E4   # <-- recommended: 2E4
mu <- 1/1000
gamma <- 365/20
R0 <- 2.5
eps <- mu/(gamma+mu)
beta <- R0*(gamma+mu)
infect.init = 1E-2

K.fct <- 'one'  
K.prm <- c(2,-5)

### ==== ODE parameters ====

timestep = 1/365
inits <- c( S = 1-infect.init, 
			I = infect.init  )
dt <- seq(0,horizon,timestep)

### ==== Stochastic parameters =====

set.seed(1234)
do.adaptivetau <- TRUE
epsilon <- 0.05  # default=0.05 ; larger=faster but less accurate
n.MC <- 9        # Monte carlo iterations

# Initial values
I0 = pop.size * infect.init   # <--- this needs to be large for cvg stochastic -> deterministic
x0 <- c(S = pop.size-I0,
		I = I0,
		R = 0)

prm.sir <- list(mu=mu, beta=beta, gama=gamma, K.fct = 'one')
prm     <- list(mu=mu, beta=beta, gama=gamma, K.fct = K.fct, K.prm = K.prm)


### ==== ABM parameters ====

params <- list(R0 = R0,
			   latent_mean = 0.00001,
			   infectious_mean = 1/gamma*365,
			   nE = 0,
			   nI = 1,
			   Kfct = K.fct, # one, sqrt, lin, exp, pow, inv
			   Kfct_prm = K.prm)

simulParams <- list(horizon = horizon*365,
					popSize = pop.size,
					seed = 1234,
					init_I1 = I0,
					calc_WIW_Re = FALSE,
					doExact = FALSE,
					timeStepTauLeap = 0.1)


### ==== ODE Simulations ====

sim.sir <- as.data.frame(lsoda(inits, dt, SIR_nlr, parms = prm.sir))
sim     <- as.data.frame(lsoda(inits, dt, SIR_nlr, parms = prm))
# Checks Simple SIR Endemic equilibrium
S.ee <- 1/R0
I.ee <- (1-1/R0)*eps


### ==== Stochastic Simulations ====

res.stoch <- run.stoch.mc(do.adaptivetau, 
						  n.MC,
						  x0,
						  transitions,
						  trans.rate,
						  params.sir = prm.sir, 
						  params = prm,
						  epsilon,
						  horizon)

sim.stoch <- res.stoch[['sim.mean']]


### ==== ABM simulations ====

sim.abm <- run_MC_ABM(params, simulParams, 
					  nMC = n.MC, 
					  nCPU = 3, 
					  libpath = libpath)
time.abm <- sim.abm[['ts']]$t / 365
prev.abm <- sim.abm[['ts']]$prev.mean


# ==== Plots ====

if(save.to.file) pdf('plot_comp_ode.pdf', width=6, height = 12)

par(mfrow=c(2,1))
lwd <- 6

plot(x= sim$time, 
	 y= sim$I * pop.size, 
	 ylim = range(sim$I * pop.size, 
	 			 sim.sir$I* pop.size,
	 			 prev.abm),
	 xlab = 'time', ylab = 'prevalence',
	 typ = 'l', las = 1,
	 lwd = lwd,
	 col = 'grey70',
	 main = 'SIR non-linear recovery\nComparison ODE vs Stochastic vs ABM')
lines(sim.sir$time, sim.sir$I* pop.size, 
	  col='grey', lty=2)

lines(sim.stoch$roundtime, sim.stoch$i.mean, col='red', lwd = lwd/3)
# lines(sim.stoch$roundtime, sim.stoch$i.q5, col='orange', lwd = 1)
# lines(sim.stoch$roundtime, sim.stoch$i.q95, col='orange', lwd = 1)

lines(x = time.abm, y = prev.abm, lwd = lwd/3, col = 'green3')


legend(x='topright', 
	   legend = c('ODE', 'Stoch', 'ABM' ,'ODE linear recov'),
	   col = c('grey70','red', 'green','grey'),
	   lwd = c(lwd, lwd/3, lwd/3, 1),
	   lty = c(1,1,1,2))

# log scale:
plot(x= sim$time, 
	 y= sim$I * pop.size, 
	 ylim = range(sim$I * pop.size, sim.sir$I* pop.size),
	 xlab = 'time', ylab = 'prevalence',
	 lwd = lwd,
	 typ = 'l', las = 1,
	 col = 'grey70',
	 log = 'y',
	 main = 'log scale')
lines(sim.sir$time, sim.sir$I* pop.size, 
	  col='grey', lty=2)
lines(sim.stoch$roundtime, sim.stoch$i.mean, col='red', lwd = lwd/3)
lines(x = time.abm, y = prev.abm, lwd = lwd/3, col = 'green3')


if(save.to.file) dev.off()

```

The prevalence curves match: the ABM has passed this basic test.


## Analysis of distributions

Let's look at the distribution of the infectious period under several $K$ recovery functions.

Let's set the main parameters for the simulation:

```{r main simulation param, echo=TRUE}
horizon         <- 365  # in days
pop.size        <- 1E4
infectious.mean <- 4.5  # in days
latent.mean     <- 3.5  # in days
R0              <- 3.00

# Number of compartments in E and I:
nE    <- 1
nI    <- 1

n.CPU <- 3           # CPU used
n.MC  <- 2 * n.CPU   # Monte carlo iterations per CPU

K.list <- list()

K.list[[1]] <- list(Kfct = 'one',
					Kfct_prm = c(0))

K.list[[2]] <- list(Kfct = 'sqrt',
					Kfct_prm = c(0))

K.list[[3]] <- list(Kfct = 'lin',
					Kfct_prm = c(0))

K.list[[4]] <- list(Kfct = 'exp',
					Kfct_prm = c(2))

K.list[[5]] <- list(Kfct = 'pow',
					Kfct_prm = c(3))

K.list[[6]] <- list(Kfct = 'inv',
					Kfct_prm = c(2, -5))

```


```{r simulation, echo=FALSE, cache=TRUE,results='hide',include=FALSE}

library(ggplot2)
library(nlr,lib.loc = libpath)
source('SIR_nlr_ABM.R')
source('nlr_utils.R')

t0 <- as.numeric(Sys.time())

set.seed(1234)
save.to.file <- FALSE

### ==== ABM parameters ====

mu        <- 1/1000
gamma     <- 1/infectious.mean
eps    <- mu/(gamma+mu)
beta   <- R0*(gamma+mu)
infect.init <- 1E-3
I0 <- pop.size * infect.init   

base.prm <- list(R0 = R0,
				 latent_mean     = latent.mean,
				 infectious_mean = infectious.mean,
				 nE = nE,
				 nI = nI)

simulParams <- list(horizon = horizon,
					popSize = pop.size,
					seed = 1234,
					init_I1 = I0,
					calc_WIW_Re = FALSE,
					doExact = FALSE,
					timeStepTauLeap = 0.1)

### ==== ABM simulations ====

# Define wrap function for parallel execution:
wrap_abm <- function(X,
					 base.prm, 
					 simulParams, 
					 nMC = n.MC, 
					 nCPU = n.CPU) {
	# Run simulation:
	params <- c(base.prm, X)
	sim.abm <- run_MC_ABM(params, 
						  simulParams, 
						  nMC = n.MC, 
						  nCPU = n.CPU,
						  libpath = libpath)
	
	# Generation interval results:
	gi.abm <- sim.abm[['gi']]
	gi.abm <- gi.abm[gi.abm$gi_bck.mean>0,]
	z <- summary.gi(gi.abm, CIwidth =0.90)
	gi.weekly.mean <- z$gi.weekly.mean
	gi.abm <- z$gi.abm
	
	# Infectiousness duration results:
	infdur.abm <- sim.abm[['infdur']]
	infdur.abm <- infdur.abm[infdur.abm$infdur.mean>0,]
	z <- summary.infdur(infdur.abm, CIwidth =0.90)
	infdur.weekly.mean <- z$infdur.weekly.mean
	infdur.abm <- z$infdur.abm
	infdur.raw <- sim.abm[['infdur.raw']]
	
	return(list(gi.abm = gi.abm,
				gi.weekly.mean = gi.weekly.mean,
				infdur.abm = infdur.abm,
				infdur.weekly.mean = infdur.weekly.mean,
				infdur.raw = infdur.raw
				)
	)
}

res <- lapply(X           = K.list, 
			  FUN         = wrap_abm,
			  base.prm    = base.prm, 
			  simulParams = simulParams, 
			  nMC         = n.MC, 
			  nCPU        = n.CPU)

df.list <- list()
dfall.list <- list()
df.infdur.list <- list()
dfall.infdur.list <- list()
df.infdur.raw <- list()

for (i in 1:length(res)) {
	# GI:
	tmp <- res[[i]]$gi.weekly.mean
	tmp$Kfct <- K.list[[i]]$Kfct
	df.list[[i]] <- tmp
	rm(tmp)
	tmp <- res[[i]]$gi.abm
	tmp$Kfct <- K.list[[i]]$Kfct
	dfall.list[[i]] <- tmp
	
	# Infectious duration:
	tmp2 <- res[[i]]$infdur.weekly.mean
	tmp2$Kfct <- K.list[[i]]$Kfct
	df.infdur.list[[i]] <- tmp2
	rm(tmp2)
	tmp2 <- res[[i]]$infdur.abm
	tmp2$Kfct <- K.list[[i]]$Kfct
	dfall.infdur.list[[i]] <- tmp2

	tmp3 <- res[[i]]$infdur.raw
	tmp3$Kfct <- K.list[[i]]$Kfct
	df.infdur.raw[[i]] <- tmp3
	rm(tmp3)
}

df    <- do.call('rbind',df.list)
dfall <- do.call('rbind',dfall.list)

df.infdur    <- do.call('rbind',df.infdur.list)
dfall.infdur <- do.call('rbind',dfall.infdur.list)
df.infdur.raw <- do.call('rbind',df.infdur.raw)

df.infdur <- subset(df.infdur, week>0)
dfall.infdur <- subset(dfall.infdur, week>0)


### ==== PLOTS ====

# if(save.to.file) pdf('plot_gi.pdf', width=12, height = 8)
# plot_mean(df = df, type = 'gi')
# plot_distribution_mean(dfall = dfall, type = 'gi')
# if(save.to.file) dev.off()
# 
# if(save.to.file) pdf('plot_infdur.pdf', width=12, height = 8)
# plot_mean(df = df.infdur, type = 'infdur')
# plot_distribution_mean(dfall = dfall.infdur, type = 'infdur')
# plot_distribution_raw(df.infdur.raw, xmax=60)
# plot_distribution_raw_time(df.infdur.raw, xmax=60)
# if(save.to.file) dev.off()
# 
t1 <- as.numeric(Sys.time())
message(paste('Time elapsed:',round((t1-t0)/60, 2), 'min'))
```


### Generation interval distributions

For the backward generation interval distributions, the mean GI seems to be increased under all transformations of the recovery process (all curves are above `one`). Note that for the "standard" case `K=one`, we find the same shape as in Champredon & Dushoff RSPB 2015.
It is also remarkable that the square root recovery streches the GIs and also the epidemic. 

```{r plot_GI}
library(ggplot2)
plot_mean(df = df, type = 'gi')

```

Looking at the evolution of GI distributions is less clear to see the differences:

```{r plot_GI_2}
plot_distribution_mean(dfall = dfall, type = 'gi')
```

### Infectiousness duration distributions

Patterns for the mean duration of infectiousness are more noisy. 

```{r plot_infdur}
plot_mean(df = df.infdur, type = 'infdur')
```

Unlike the GIs, it is easier to see what's going on by looking directly at the infectious duration distributions directly. Note the black dashed line is the exponential density, which must match the `K=one` case when `nI=1`:
```{r plot_infdur_distribution}
plot_distribution_raw(df.infdur.raw, xmax=60)
```

There seems to be a remarkable point (around infdur=6) where the shape for $K$ does not affect the density. Can we find it analitycally???

Now, looking at how these distributions evolve with calendar time: 

```{r}
plot_distribution_raw_time(df.infdur.raw, xmax=60)
```

It is not clear because of the noise from the weeks where there is no/very little data.
